A Triangle Mesh Approach to Painting

### 2014-04-21 template.tex
Added new figures to make it look pretty :D

\begin{abstract}
Vector graphics have a compact representation, are easy to manipulate, and have infinite resolution.
Traditionally, vector graphics have required artists to master a new skill in order to effectively utilize
them. Recent work has converted raster graphics to vector graphics, allowed multi-resolution raster painting,
and created new ways to represent vector graphics. However, none of these have solved the problem of
letting an artist use standard and intuitive raster painting techniques to generate vector graphics.
Our algorithm, TrianglePainter, converts a user's strokes into a triangle mesh vector representation. 
TrianglePainter lets artists focus on the art, while it handles the vectorization in the background.

@@

This representation allows painting at any scale. A user can make large strokes
while zoomed out and then zoom in to make fine details. All of this can happen without
loss of quality since the data is stored using triangles instead of pixels, which can represent sharp boundaries at any orientation.  Furthermore, the use of smooth shading when rendering the triangles allows for vector representation of gradient effects.  We take advantage of this feature to create soft, airbrush-like strokes, which are usually very difficult to create in vector drawing software.

@@

\begin{figure}
    \centering
        \includegraphics[width=0.4\textwidth]{images/colorarcs}
    \caption{An example of color arcs. The orange lines are boundary edges. Black lines form the triangle
    mesh with the boundary edges. Each point is given a color arc which defines its color in that range.
    Note that the points in the middle row have different colors on top and bottom. This gives blue
    triangles on top and red on bottom with a hard boundary in between.}
    \label{fig:arcs}
\end{figure}

@@

Since the conditions on color arcs make colors emanate from boundaries it is easy to create regions (e.g. a brush stroke boundary)
containing either a solid color or linear gradient. A region will have a solid color (e.g. a hard brush stroke) if the
boundaries that surround it all take on the same color. If instead a region is surrounded by
two boundaries, one which has a given color and the other which has a transparent version of that color
then there will be a linear gradient in that region between the two boundaries (e.g. a soft brush stroke).

\begin{figure}
    \centering
        \includegraphics[height=1.5in]{images/softstrokeboundaryedges}
    \caption{Place Holder.}
\end{figure}

@@

\begin{figure*}
\centering
\begin{subfigure}[b]{0.3\textwidth}
  \centering
  \includegraphics[width=.9\textwidth]{images/stroke_triangulation/hardrendered}
  \caption{The rasterized stroke.}
  \label{fig:sub1}
\end{subfigure}%
\begin{subfigure}[b]{0.3\textwidth}
  \centering
  \includegraphics[width=.9\textwidth]{images/stroke_triangulation/hardpruned}
  \caption{The contour of the stroke.}
  \label{fig:sub2}
\end{subfigure}
\begin{subfigure}[b]{0.3\textwidth}
  \centering
  \includegraphics[width=.9\textwidth]{images/stroke_triangulation/hardmesh}
  \caption{The triangulation of the contour}
  \label{fig:sub2}
\end{subfigure}
\caption{Triangulating a hard stroke.}
\label{fig:test}
\end{figure*}

\begin{figure*}
\centering
\begin{subfigure}[b]{0.3\textwidth}
  \centering
  \includegraphics[width=.9\textwidth]{images/stroke_triangulation/softrendered}
  \caption{The rasterized stroke.}
  \label{fig:sub1}
\end{subfigure}%
\begin{subfigure}[b]{0.3\textwidth}
  \centering
  \includegraphics[width=.9\textwidth]{images/stroke_triangulation/softpruned}
  \caption{The contour of the stroke.}
  \label{fig:sub2}
\end{subfigure}
\begin{subfigure}[b]{0.3\textwidth}
  \centering
  \includegraphics[width=.9\textwidth]{images/stroke_triangulation/softmesh}
  \caption{The triangulation of the contour}
  \label{fig:sub2}
\end{subfigure}
\caption{Triangulating a soft stroke.}
\label{fig:test}
\end{figure*}

@@

Again, using marching squares these contours are extracted and then pruned. The boundaries and points
derived from the contours are given to Triangle and a mesh is returned. This mesh represents the soft
stroke where no triangle crosses the boundary between the soft and hard regions
of the stroke. Points on the contour of the inner stroke will have color arcs with the stroke color,
while the points on the contour of the outer stroke will have transparent color arcs. This gives
all triangles connecting the inner and outer strokes a linear gradient (see Figure~\ref{XXX}).

\begin{figure}
    \centering
        \includegraphics[height=1.5in]{images/softstroke}
        \includegraphics[height=1.5in]{images/hardvssoft}
    \caption{Place Holder.}
\end{figure}

@@

In the simplest approach, the points and boundaries from the new stroke are added
to the list of points and boundaries already stored on the canvas. All of this data is given to Triangle
to produce a new mesh. The new mesh contains all of the points of the old and new strokes. Since
all of the boundaries are stored and given to Triangle, the stroke edges are preserved. This means that
no triangle can cross the boundary of a stroke. Without this constraint, it would be impossible to correctly color the
triangles since they would straddle color arcs. Once a new mesh is formed the color arcs of the 
new and old points must be computed.

\begin{figure}
    \centering
        \includegraphics[height=1.5in]{images/canvastriangulation}
    \caption{Place Holder.}
\end{figure}

@@

An intersection point lies on two boundaries and therefore has four boundary edges emanating from it. 
To determine the colors of an intersection points requires knowing the colors associated with
other points on the boundaries. For both boundaries, travel both directions to the first point with color data. Note this may not
be the first point on each side since this intersection point may be adjacent to other intersection
points also without color data. For both boundaries, the colors in the color arcs are linearly interpolated. 
This yields four average colors, two from the new stroke boundary and two from the old stroke boundary.
The colors from the new stroke are composited over the colors from the old stroke to yield four new
colors, one for each region defined by the boundary intersections.

\begin{figure}
    \centering
        \includegraphics[height=1.5in]{images/determiningcolors}
    \caption{Place Holder.}
\end{figure}

@@

\begin{figure}
    \centering
        \includegraphics[height=1.5in]{images/findingmodifiedpoints}
    \caption{Place Holder.}
    \label{fig:modified}
\end{figure}

\section{Discussion and Results}
Paintings made with TrianglePainter can be seen in Figures~\ref{fig:face} and \ref{fig:boxes}.  Here we first consider how the complexity of the algorithm grows both in time and geometry.
The results from this analysis and our experience with TrianglePainter have lead to a number of observations.

@@

Figure~\ref{fig:numtriangles} shows the zooming case grows linearly while the non-zooming case
appears to grow quadratically. This is expected since every new stroke has the chance to overlap
all of the old strokes. This means each new stroke does not just add a constant amount of geometry.
Due to intersections with previous strokes intersection points are added. This results in a linear
increase in the amount of geometry. This results
in a quadratic growth rate. However, in the zooming case each stroke rarely overlaps and therefore
intersection points are rarely added. Instead the geometry increases by a constant which leads to
a linear growth rate.

@@

\subsection{Performance}
The current implementation uses Python and PyOpenGL. Python's performance is slower than
a lower level implementation in C++ would be. The drawing of the canvas is quick with
no lag. However, the actual process to take a stroke from rasterization to triangle
representation on the canvas can be slow. On a blank canvas, strokes take less than
one second to process. However, on a canvas with lots of geometry a stroke that overlaps
that geometry can take several seconds to render. Some of this could be alleviated by a carefully optimized implementation. However, there are many calculations
necessary for the old points affected by the new stroke and the points associated with the
new stroke itself. Since these geometries can become arbitrarily complicated (in pathological cases), these
calculations can take an arbitrarily long. This brings us to the final point, scalability.

\subsection{Simplification}
No matter how good the implementation, the fact that paintings can become arbitrarily complicated as more and more strokes are added
means there must be some way to simplify the mesh. In most situations where
many strokes overlap one another, most of the geometry is redundant. For example, when an opaque stroke is laid down on top of a complex geometry, that complex
geometry no longer has any useful information since the outline of the opaque stroke
completely defines it (all vertices are the same color).  Furthermore, even with translucent or feathered strokes, a large enough number of 
composited strokes may produce complex geometry that does not add much to the visual appearance of the canvas. Some of this geometry can be reduced without noticeable changes.  To accomplish this, our proposed simplification algorithm would identify a vertex that can be removed via an edge collapse by measuring the change the collapse would cause in the image.  The change is computed as the magnitude of the color shift and the translation of a boundary.  If these deltas are below a threshold, then the collapse can proceed.  This process could be done when a stroke is added to the canvas, or continually in a background thread to avoid increasing apparent latency.  Though we have yet to implement this simplification algorithm, it is important to ensure a painter can continue iterate on a painting without the program becoming too slow.

@@

>>>>

\begin{abstract}
Although vector graphics program have a number of advantages, current vector painting programs
do not allow an artist to use traditional painting techniques. We propose a new algorithm that translates a
user's mouse motion into a triangle mesh representation. This triangle mesh can then be composited onto
a canvas containing an existing mesh representation of the previous strokes. 
This representation allows the algorithm to render solid colors and linear
gradients. It also enables painting at any resolution. This gives artists the opportunity to create complex,
multi-scale drawings with gradients while avoiding artifacts.

@@

This procedure allows painting at any scale. A user can make large strokes
while zoomed out and then zoom in to make fine details. All of this can happen without
loss of quality since the data is stored using triangles instead of pixels, which can represent sharp boundaries at any orientation.  Furthermore, the use of smooth shading when rendering the triangles allows for vector representation of gradient effects.  We take advantage of this feature to create soft, airbrush-like strokes, which are usually very difficult to create in vector drawing software.

@@

\begin{figure}
    \centering
        \includegraphics[width=0.4\textwidth]{images/colorarcs2}
    \caption{An example of color arcs. The orange lines are boundary edges. Black lines form the triangle
    mesh with the boundary edges. Each point is given a color arc which defines its color in that range.
    Note that the points in the middle row have different colors on top and bottom. This gives blue
    triangles on top and red on bottom with a hard boundary in between.}
    \label{fig:arcs}
\end{figure}

@@

Since the conditions on color arcs make colors emanate from boundaries it is easy to create regions (e.g. a brush stroke boundary)
containing either a solid color or linear gradient. A region will have a solid color (e.g. a hard brush stroke) if the
boundaries that surround it all take on the same color. If instead a region is surrounded by
two boundaries, one which has a given color and the other which has a transparent version of that color
then there will be a linear gradient in that region between the two boundaries (e.g. a soft brush stroke).
Figure~\ref{fig:softarcs} shows the color arcs on such a soft stroke.

\begin{figure}
    \centering
        \includegraphics[width=0.5\textwidth]{images/gradientarcs}
    \caption{A close look at a gradient stroke. The blue edges are boundaries where
    the points have gray color arcs as shown. The purple edges are boundaries where
    the points have transparent gray color arcs. This gives triangles connecting the two boundaries
    a linear gradient from gray to white.}
    \label{fig:softarcs}
\end{figure}

@@

\begin{figure}
    \centering
        \includegraphics[width=.2\textwidth]{images/stroke_triangulation/hardrendered}
        \includegraphics[width=.2\textwidth]{images/stroke_triangulation/softrendered}

        \includegraphics[width=.2\textwidth]{images/stroke_triangulation/hardpruned}
        \includegraphics[width=.2\textwidth]{images/stroke_triangulation/softpruned}

        \includegraphics[width=.2\textwidth]{images/stroke_triangulation/hardmesh}
        \includegraphics[width=.2\textwidth]{images/stroke_triangulation/softmesh}
    \caption{Left: A hard stroke is rasterized, the contour is found, and a triangle mesh is generated.
             Right: A soft stroke is rasterized in two different shades of gray, two contours are found,
             a triangle mesh is generated with a boundary separating the inner and outer parts of the stroke.}
    \label{fig:rastertotriangles}
\end{figure}

@@

Again, using marching squares these contours are extracted and then pruned. The boundaries and points
derived from the contours are given to Triangle and a mesh is returned. This mesh represents the soft
stroke where no triangle crosses the boundary between the soft and hard regions
of the stroke. Points on the contour of the inner stroke will have color arcs with the stroke color,
while the points on the contour of the outer stroke will have transparent color arcs. This gives
all triangles connecting the inner and outer strokes a linear gradient (again see Figure~\ref{softarcs}).
Figure~\ref{fig:rastertotriangles} shows the process of triangulating both hard and soft strokes.

% \begin{figure}
%     \centering
%         \includegraphics[height=1.5in]{images/softstroke}
%         \includegraphics[height=1.5in]{images/hardvssoft}
%     \caption{Place Holder.}
% \end{figure}

@@

In the simplest approach, the points and boundaries from the new stroke are added
to the list of points and boundaries already stored on the canvas. All of this data is given to Triangle
to produce a new mesh (see figure~\ref{fig:firststrokes}). The new mesh contains all of the points of the old and new strokes. Since
all of the boundaries are stored and given to Triangle, the stroke edges are preserved. This means that
no triangle can cross the boundary of a stroke. Without this constraint, it would be impossible to correctly color the
triangles since they would straddle color arcs. Once a new mesh is formed the color arcs of the 
new and old points must be computed.

\begin{figure}
    \centering
        \includegraphics[width=.23\textwidth]{images/emptycanvas}
        \includegraphics[width=.23\textwidth]{images/canvaswithonestroke}
    \caption{The figure on the left shows an empty canvas with a number of points and boundaries defined along the edges
             to provide the initial triangulation. The figure on the right shows how a stroke is composited onto
             the empty canvas. Blue edges are boundary edges. Red edges form the triangle mesh with the blue edges.}
    \label{fig:firststrokes}
\end{figure}

@@

An intersection point lies on two boundaries and therefore has four boundary edges emanating from it. 
To determine the colors of an intersection points requires knowing the colors associated with
other points on the boundaries. For both boundaries, travel both directions to the first point with color data. Note this may not
be the first point on each side since this intersection point may be adjacent to other intersection
points also without color data. For both boundaries, the colors in the color arcs are linearly interpolated. 
This yields four average colors, two from the new stroke boundary and two from the old stroke boundary.
The colors from the new stroke are composited over the colors from the old stroke to yield four new
colors, one for each region defined by the boundary intersections (see figure~\ref{fig:intersectioncolors})

An example of how color arcs are determined for both intersection and regular points see figure~\ref{fig:colorarcdeterminer}.

\begin{figure}
    \centering
        \includegraphics[width=.2\textwidth]{images/intersectioncolorarcs1new}
        \includegraphics[width=.2\textwidth]{images/intersectioncolorarcs2new}

    \caption{Left: A blue stroke is composited over a red stroke. Point A is covered
             by the the blue stroke and point B is on top of the red stroke. Middle:
             New color arcs are determined. Both of A's color arcs are composited
             with the blue stroke which turns red to purple and transparent to blue. Similarly
             both of B's color arcs are composited on top of the red stroke which turns
             blue to purple and transparent to red. Two intersection points C and D are
             added. Their color arcs are determined by examining neighboring points. Right: 
             A new triangle mesh is generated to preserve boundaries. Note that triangles
             in the intersection will be purple.}
    \label{fig:intersectioncolors}
\end{figure}

\begin{figure*}
    \centering
        \includegraphics[width=.32\textwidth]{images/composite1new}
        \includegraphics[width=.32\textwidth]{images/composite2new}
        \includegraphics[width=.32\textwidth]{images/composite3new}
    \caption{Left: A blue stroke is composited over a red stroke. Point A is covered
             by the the blue stroke and point B is on top of the red stroke. Middle:
             New color arcs are determined. Both of A's color arcs are composited
             with the blue stroke which turns red to purple and transparent to blue. Similarly
             both of B's color arcs are composited on top of the red stroke which turns
             blue to purple and transparent to red. Two intersection points C and D are
             added. Their color arcs are determined by examining neighboring points. Right: 
             A new triangle mesh is generated to preserve boundaries. Note that triangles
             in the intersection will be purple.}
    \label{fig:colorarcdeterminer}
\end{figure*}

@@

\begin{figure}
    \centering
        \includegraphics[width=.5\textwidth]{images/grid}
    \caption{An example of how TrianglePainter determines which triangles to save and which to retriangulate. Blue
    lines represent the boundaries of old strokes, orange lines are the boundaries of the new stroke, red lines are
    the triangles that will be reused, and purple lines define the area that needs to be retriangulated. The purple
    lines are edges of triangles that intersect the new stroke. Since they interesect the new stroke they are no
    longer valid and must be retriangulated. All other triangles on the canvas can be reused.}
    \label{fig:modified}
\end{figure}

\section{Discussion and Results}
Paintings made with TrianglePainter can be seen in Figures~\ref{fig:face}, \ref{fig:rainbow}, \ref{fig:stop} and \ref{fig:boxes}.  Here we first consider how the complexity of the algorithm grows both in time and geometry.
The results from this analysis and our experience with TrianglePainter have lead to a number of observations.

@@

Figure~\ref{fig:numtriangles} shows the zooming case grows linearly while the non-zooming case
appears to grow quadratically. This is expected since every new stroke has the chance to overlap
all of the old strokes. Therefore each new stroke does not just add a constant amount of geometry, but 
due to intersections with previous strokes extra intersection points are added. This results in a linear
increase in the amount of geometry per stroke yielding a quadratic growth rate. However, in the zooming case each stroke rarely overlaps and therefore
intersection points are rarely added. In this case the geometry increases by a constant amount per stroke which yields
a linear growth rate.

@@

\subsection{Performance}
The current implementation uses Python and PyOpenGL. Python's performance is slower than
a lower level implementation in C++ would be. The drawing of the canvas is quick with
no lag. However, the actual process to take a stroke from rasterization to triangle
representation on the canvas can be slow. On a blank canvas, strokes take less than
one second to process. However, on a canvas with lots of geometry a stroke that overlaps
that geometry can take several seconds to render. Some of this could be alleviated by a carefully optimized implementation. However, there are many calculations
necessary for the old points affected by the new stroke and the points associated with the
new stroke itself. Since these geometries can become arbitrarily complicated (in pathological cases), these
calculations can take an arbitrarily long. This brings us to the final point, simplification.

\subsection{Simplification}
No matter how good the implementation, the fact that paintings can become arbitrarily complicated as more and more strokes are added
means there must be some way to simplify the mesh. In most situations where
many strokes overlap one another, most of the geometry is redundant. For example, when an opaque stroke is laid down on top of a complex geometry, that complex
geometry no longer has any useful information since the outline of the opaque stroke
completely defines it (all vertices are the same color).  Furthermore, even with translucent or feathered strokes, a large enough number of 
composited strokes may produce complex geometry that does not add much to the visual appearance of the canvas. Some of this geometry can be reduced without noticeable changes.  To accomplish this, our proposed simplification algorithm would identify a vertex that can be removed via an edge collapse by measuring the change the collapse would cause in the image.  The change is computed as the magnitude of the color shift and the translation of a boundary.  If these deltas are below a threshold, then the collapse can proceed.  This process could be done when a stroke is added to the canvas, or continually in a background thread to avoid increasing apparent latency.  Though we have yet to implement this simplification algorithm, it is important to ensure a painter can continue to iterate on a painting without the program becoming too slow.

@@

\begin{figure*}
    \centering
        \includegraphics[width=0.49\textwidth]{images/rainbow2}
        % \includegraphics[width=0.49\textwidth]{images/facezoomtriangles}
    \caption{An example of soft strokes}
    \label{fig:rainbow}
\end{figure*}

\begin{figure*}
    \centering
        \includegraphics[width=0.3\textwidth]{images/stop1}
        \includegraphics[width=0.3\textwidth]{images/stop2}
        \includegraphics[width=0.3\textwidth]{images/stop3}
    \caption{}
    \label{fig:stop}
\end{figure*}
